\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2017-2018

### 2017-09-19 auspice/reports/tex/sep-2017.tex
More revisions

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) is beginning to decline in frequency and
is now at $\sim$21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

@@

The most notable subclade of 3c3.a is a clade carrying the mutation
F193S. However, this clade along with its sister clades has been
decreasing in recent month and it doesn't show a consistent signal of
antigenic change. The F193D mutation is observed sporadically within
the 3c2.a clade as well.

% \begin{figure}[h!]
%   \centering
%   \includegraphics[width=1.0\textwidth]{../figures/sep-2017/h3n2_3c2a.png}
%   \caption{\textbf{Recent frequency changes.} The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 4 month.
%   }
%   \label{H3N2_135K}
% \end{figure}

@@

The S74R and I295V clade has risen rapidly in frequency starting around
Jan 2017 (\FIG{h1n1pdm_mutations}). This clade has been very successful and is now at
approximately 92\% global frequency. Within this clade the subclade
comprising S164T has risen in frequency from April 2017 and is now at
approximately 35\% global frequency.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h1n1pdm_mutations_6b_6b1.png}
  \caption{\textbf{Sequential rapid replacement of clades in A/H1N1pdm.} The latest rapid rise of 74R took place at similar or greater speed as the spread of 162N.
  }
  \label{h1n1pdm_mutations_6b_6b1}
\end{figure}
It was previously observed that clade 6b viruses with mutation K163Q
showed decreased titers in a subset of adult human sera due to
immunodominance effects, despite continued high titers in ferret sera
\cite{linderman2014potential}.
Importantly, the new S74R and I295V clade has spread at a similarly
rapid rate as previous clades 6b and 6b.1 (\FIG{h1n1pdm_mutations_6b_6b1}). This rate of spread and
displacement of existing viral diversity is consistent with a selective
origin.

\emph{Given current patterns of growth and decline it is almost certain
that the S74R and I295V clade will dominant in a years time. Although
there is no antigenic effect visible via HI from ferret antisera, the
rapidity of the rise of S74R and I295V suggests a selective origin. We
would suggest a detailed examination of human serological data as it's
already been established that human serology may differ from ferret
serology in H1N1pdm viruses.}

>>>>

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

@@

The most notable subclade of 3c3.a is a clade carrying the mutation
F193S. However, this clade along with its sister clades has been
decreasing in recent month and it doesn't show a consistent signal of
antigenic change. The F193S mutation is observed sporadically within
the 3c2.a clade as well.

Measuring clade frequency growth and decline over the previous 6 months shows that subclades within clades 3 and 5 are growing the most rapidly (\FIG{h3n2_tree_dfreq}).
The most rapidly growing subclade is the E62G/R142G/T135K subclade within clade 5.
This analysis is borne out in "local branching index" \cite{neher2014predicting}, which also supports this subclade as a high fitness clade.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_dfreq.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by clade growth.}
  The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 6 months.
  }
  \label{h3n2_tree_dfreq}
\end{figure}

@@

The S74R and I295V clade has risen rapidly in frequency starting around Jan 2017 (\FIG{h1n1pdm_mutations}).
This clade has been very successful and is now at approximately 66\% global frequency.
Notably, this clade is at \textgreater{}90\% frequency in North America, Europe and Oceania but is at lower frequency in Asia.
Within this clade the subclade comprising S164T has risen in frequency from April 2017 and is now at approximately 32\% global frequency.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h1n1pdm_mutations_6b_6b1.png}
  \caption{\textbf{Sequential rapid replacement of clades in A/H1N1pdm.} The latest rapid rise of 74R took place at similar or greater speed as the spread of 162N.
  }
  \label{h1n1pdm_mutations_6b_6b1}
\end{figure}
It was previously observed that clade 6b viruses with mutation K163Q
showed decreased titers in a subset of adult human sera due to
immunodominance effects, despite continued high titers in ferret sera
\cite{linderman2014potential}.
Importantly, the new S74R and I295V clade has spread at a similar
rate as previous clades 6b and 6b.1 (\FIG{h1n1pdm_mutations_6b_6b1}). This rate of spread and
displacement of existing viral diversity is consistent with a selective
origin.

\emph{Given current patterns of growth and decline it is highly likely
that the S74R and I295V clade will dominant in a years time. Although
there is no antigenic effect visible via HI from ferret antisera, the
rapidity of the rise of S74R and I295V suggests a selective origin. The S164T subclade also appears to be selectively driven. We
would suggest a detailed examination of human serological data as it's
already been established that human serology may differ from ferret
serology in H1N1pdm viruses.}
